BOLD Parameter Estimation Using Particle Filters

### 2010-07-21 doc/thesis/background.tex
corrected the intro, and moved the noise description out to methods

\chapter{Prior Works}
\label{sec:Prior Works}
The ultimate purpose of this work is to provide a new tool
for analyzing FMRI data.

@@

\subsection{Classical Activation Detection}
\label{sec:Square Wave}
The most basic method of analyzing FMRI data is through a standard T-test
between "resting state" and "active state" samples. Simply put, the
mean is calculated separately for non-stimulus and stimulus time intervals.
A classic t-test may then be applied, giving the probability that the
distributions actually have the same mean. Because of the correlated
noise present in FMRI (\autoref{sec:Introduction Noise}), it is necessary
to apply some sort of high-pass filter to the data. Without applying
such a filter, P values must be set extraordinarily high to prevent
false positives \cite{Smith2007}. If there truly is signal
due to stimuli, the distributions will not actually be independent
Gaussians because activation does not fit a square wave 
(\autoref{sec:BOLD Physiology}). For this reason other methods
are often more used, as discussed in 
\autoref{sec:Current Techniques General Linear Model}.

@@

As mentioned previously, a square wave stimulus 
does not result in a square wave in the activation of brain regions. 
The BOLD signal is in fact a significantly smoothed version of the 
stimuli. As such, when fitting an FMRI time course to the input,
the input ($X$'s columns) is usually smoothed to reduce bias error. 
The best method, that maintains a linear fit, is convolving the input 
with a Hemodynamic 
Response Function (HRF). The so called "Hemodynamic Response Function"
mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The fitting 
process is then a least squares fit over the space of the vector $\beta$. 
Therefore $Y(t)$ is estimated as a linear combination 
of the columns of $X$.

Smoothing the input with a single HRF poses certain problems.
It is well known that different Hemodynamic Response Functions are necessary 
for different regions of the brain. The "Canonical" HRF that is most used, 
has been optimized for the visual cortex. As \autoref{sec:Post Stimulus Undershoot}
discussed, there are certainly variations in the shape of the
BOLD response, both between brain regions and patients.  \cite{Handwerker2004}
discusses the implications of choosing an incorrect HRF, the most important
of which is a definite increase in false negatives. While an atlas
of Hemodynamic Response Functions for each region could definitely 
mitigate this risk, it does not deal with variation between patients.
Thus, the inability to fit parameters other than scale definitely hinders 
analysis of activation. As a result, there is significant bias error
when using a linear fit to localize activation. Notably, studies
will also have a bias toward the visual cortex, precisely because
it is so well studied.

@@

Another potential method of interest is the use of Genetic Algorithms
(GA). Genetic algorithms are similar to Simulated Annealing, in
that they randomly move to better solutions based on a cost function.
However; in genetic algorithms a single point estimate isn't used. Instead
a population of estimates is generated, each with distinct parameters,
and then each set of parameters is rated with a fitness function. Parameter
sets that are good get a higher "fitness"; then new parameter sets are generated by 
randomly combining pieces of the old parameter sets. The pieces are typically
chosen at a rate proportional to the fitness of the donor; thus "fit"
parameter sets tend to pass on their properties. In addition to this,
random mutations may be introduced that come from no existing parent. 
The new generation is then rated with the fitness function again, and
the entire process starts over. The stop condition for a genetic algorithm
is typically based on some threshold for fitness or a maximum number 
of generations. This is problematic, since it doesn't really provide 
any guarantee of even reaching a local minima, although in practice
it can be quite effective.

@@

In order to estimate the posterior at $t$, a deterministic set of "sigma" points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004}, \cite{Hu2009} mention a Jacobian of the BOLD response, this is
not strictly the case and is rather $\frac{\partial J}{\partial t}$
rather than a true Jacobian.  This is important because the Extended
Kalman filter depends on the Jacobian to map a Gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does. In fact
\cite{Hu2009} uses the UKF to perform a similar type of analysis to
the one performed in this work.

>>>>

\chapter{Methods of Analyzing FMRI}
\label{sec:Prior Works}
The ultimate purpose of this work is to provide a new tool
for analyzing FMRI data.

@@

\subsection{Classical Activation Detection}
\label{sec:Square Wave}
The most basic method of analyzing FMRI data is through a standard T-test
between resting state and active state samples. Simply put, the
mean is calculated separately for non-stimulus and stimulus time intervals.
A classic t-test may then be applied, giving the probability that the
distributions actually have the same mean. Because of the correlated
noise present in FMRI (\autoref{sec:Introduction Noise}), it is necessary
to apply some sort of high-pass filter to the data. Without applying
such a filter, P values must be set extraordinarily high to prevent
false positives \cite{Smith2007}. If there truly is signal
due to stimuli, the distributions will not actually be independent
Gaussians because activation does not fit a square wave 
(\autoref{sec:BOLD Physiology}). For this reason other methods
are often more used, as discussed in 
\autoref{sec:Current Techniques General Linear Model}.

@@

As mentioned previously, a square wave stimulus 
does not result in a square wave in the activation of brain regions. 
The BOLD signal is in fact a significantly smoothed version of the 
stimuli. As such, when fitting an FMRI time course to the input,
the input ($X$'s columns) is usually smoothed to reduce bias error. 
The best method, that maintains a linear fit, is convolving the input 
with a Hemodynamic 
Response Function (HRF). The \emph{Hemodynamic Response Function}
mimics the basic shape of BOLD activation, including a delay
due to rise time and fall time. The fitting 
process is then a least squares fit over the space of the vector $\beta$. 
Therefore $Y(t)$ is estimated as a linear combination 
of the columns of $X$.

Smoothing the input with a single HRF poses certain problems.
It is well known that different Hemodynamic Response Functions are necessary 
for different regions of the brain. The Canonical HRF that is most used, 
has been optimized for the visual cortex. As \autoref{sec:Post Stimulus Undershoot}
discussed, there are certainly variations in the shape of the
BOLD response, both between brain regions and patients.  \cite{Handwerker2004}
discusses the implications of choosing an incorrect HRF, the most important
of which is a definite increase in false negatives. While an atlas
of Hemodynamic Response Functions for each region could definitely 
mitigate this risk, it does not deal with variation between patients.
Thus, the inability to fit parameters other than scale definitely hinders 
analysis of activation. As a result, there is significant bias error
when using a linear fit to localize activation. Notably, studies
will also have a bias toward the visual cortex, precisely because
it is so well studied.

@@

Another potential method of interest is the use of Genetic Algorithms
(GA). Genetic algorithms are similar to Simulated Annealing, in
that they randomly move to better solutions based on a cost function.
However; in genetic algorithms a single point estimate isn't used. Instead
a population of estimates is generated, each with distinct parameters,
and then each set of parameters is rated with a fitness function. Parameter
sets that are good get a higher weight; then new parameter sets are generated by 
randomly combining pieces of the old parameter sets. The pieces are typically
chosen at a rate proportional to the fitness of the donor; thus fit
parameter sets tend to pass on their properties. In addition to this,
random mutations may be introduced that come from no existing parent. 
The new generation is then rated with the fitness function again, and
the entire process starts over. The stop condition for a genetic algorithm
is typically based on some threshold for fitness or a maximum number 
of generations. This is problematic, since it doesn't really provide 
any guarantee of even reaching a local minima, although in practice
it can be quite effective.

@@

In order to estimate the posterior at $t$, a deterministic set of sigma points 
(often 2 per dimension, plus 1 at the mode of the multivariate distribution)
weighted according to a Gaussian estimate of $X(t-1)$ are passed through
the update equation. This set of points are then used to estimate the 
mean and covariance of $X(t)$. The benefit of this, is that it requires
no Jacobian and only a few extra calculations to get a decent esimate of
a posterior Gaussian. In the BOLD case, the set of equations we are modeling have no 
closed form solution, and finding the Jacobian is impossible without approximations. Although 
\cite{Riera2004}, \cite{Hu2009} mention a Jacobian of the BOLD response, this is
not strictly the case and is rather $\frac{\partial J}{\partial t}$
rather than a true Jacobian.  This is important because the Extended
Kalman filter depends on the Jacobian to map a Gaussian through the 
advancement of time. Thus the popular Extended Kalman Filter won't work
in this case, whereas the Unscented Kalman Filter still does. In fact
\cite{Hu2009} uses the UKF to perform a similar type of analysis to
the one performed in this work.

### 2010-07-21 doc/thesis/conclusion.tex
corrected the intro, and moved the noise description out to methods

\chapter{Conclusion}
In this paper I have proposed and demonstrated the use of a particle filter
for the estimation of BOLD model parameters. In doing so, it has been possible
to estimate the BOLD time-series and judge the quality of the estimated time-
series with respect to the true BOLD signal. As result, it is possible to use
the particle filter to localize areas where a known stimuli most directly drives
neural activation. In the future it will be possible to use the estimated states
($v,q,s,f$) to drive other models and learn more about how regions of the brain
interact.

A limitation often reached in previous works was an inconsistency of
parameter estimates, most likely because of covariance between the model parameters.
Although individual studies got consistent results, those results often differed
widely from other similar studies. The reason for this is rather clear from the
simulation results in \autoref{sec:SimLowNoise}. There is a significant amount of
trade off between parameters to the point that a signal set of parameters
is most likely not possible to derive from the BOLD response alone. It
will therefore be beneficial to combine BOLD studies with cerebral blood
flow or cerebral blood volume studies to gain multiple more measurements and
further constrain the model. That said, the benefit of the particle filter
is that it provides a full posterior distribution at the final time step.
As such the true solution should be encoded in the particle filter's final 
distribution given priors that encompass the true parameters. This is beneficial
in two ways; first, if, after the fact, some parameter becomes known 
from outside observation, it is then possible to construct a new probability
conditional on the new observation. Secondly the results from multiple runs
may be reasonably compiled together, using the final distribution from the previous
run as the prior distribution of the next run. Rather than simply providing a 
staring point for parameters to converge from, it in fact continues convergence
from the previous stopping point.

Although many versions of the BOLD model exist, and it is tempting to use 
more detailed models; from the results found here the issue of bias error
from the BOLD model is not the biggest concern. Clearly reducing variance
in parameters should be the first concern. 
One future study that would be extremely beneficial in this way  would
be an extensive study of what the priors should truly be. Although \cite{Friston2000}
gives an estimate of what is thought to be "reasonable" values, and later studies
published their estimate of the distributions, given the interplay between
parameters it is very unlikely that these priors are true to actual distribution
that occurs in vivo. As I mentioned previously, the addition of simultaneous
flow or volume measurements are another potentially powerful way to further confine the model,
and thus deal with the elasticity of parameters. As I mentioned in 
\autoref{sec:BackgroundConclusion}, a chief advantage of using physiologically
plausible models is that such data may in fact be added with relative ease.

>>>>

\chapter{Conclusion}
\label{sec:Conclusion}
In this paper I have proposed and demonstrated the use of a particle filter
for the estimation of BOLD model parameters. In doing so, it has been possible
to estimate the BOLD time-series and judge the quality of the estimated time-
series with respect to the true BOLD signal. As result, it is possible to use
the particle filter to localize areas where a known stimuli most directly drives
neural activation. In the future it will be possible to use the estimated states
($v,q,s,f$) to drive other models and learn more about how regions of the brain
interact.

A limitation often reached in previous works was an inconsistency of
parameter estimates, most likely because of covariance between the model parameters.
Although individual studies got consistent results, those results often differed
widely from other similar studies. The reason for this is rather clear from the
simulation results in \autoref{sec:SimLowNoise}. There is a significant amount of
trade off between parameters to the point that a signal set of parameters
is most likely not possible to derive from the BOLD response alone. It
will therefore be beneficial to combine BOLD studies with cerebral blood
flow or cerebral blood volume studies to gain multiple more measurements and
further constrain the model. That said, the benefit of the particle filter
is that it provides a full posterior distribution at the final time step.
As such the true solution should be encoded in the particle filter's final 
distribution given priors that encompass the true parameters. This is beneficial
in two ways; first, if, after the fact, some parameter becomes known 
from outside observation, it is then possible to construct a new probability
conditional on the new observation. Secondly the results from multiple runs
may be reasonably concatenated, using the final distribution from the previous
run as the prior distribution of the next run. Rather than simply providing a 
staring point for parameters to converge from, it in fact continues convergence
from the previous stopping point.

Although many versions of the BOLD model exist, and it is tempting to use 
more detailed models; from the results found here the issue of bias error
from the BOLD model is not the biggest concern. Clearly clarifying the distributions
of parameters for the prior should be the first concern. 
One future study that would be extremely beneficial in this way  would
be an extensive study of what the priors should truly be. Although \cite{Friston2000}
gives an estimate of what is thought to be reasonable values, and later studies
published their estimate of the distributions, given the interplay between
parameters it is very unlikely that these priors are true to actual distribution
that occurs in vivo. As I mentioned previously, the addition of simultaneous
flow or volume measurements are another potentially powerful way to further confine the model,
and thus deal with the elasticity of parameters. As I mentioned in 
\autoref{sec:BackgroundConclusion}, a chief advantage of using physiologically
plausible models is that such data may in fact be added with relative ease.

### 2010-07-21 doc/thesis/intro.tex
corrected the intro, and moved the noise description out to methods

\chapter{Introduction}
Traditional methods of analyzing images produced by Function Magnetic Resonance 
Imaging (FMRI) use regression techniques with very few degrees of freedom. 
Though adding more degrees of freedom naturally mandates more computation,
in this thesis I will discuss a Sequential Monte Carlo method of fitting a nonlinear
model with seven degrees of freedom at 
a computation cost that would allow real time calculations for a multiple voxels
(4 dimensional pixels). More practically, the method described here forms a completely
separate method of detecting neural activity in comparison to the traditionally
used Statistical Parametric Mapping (SPM). Though more computationally intense,
this method is capable of modeling nonlinear effects, is conceptually simpler
and provides more detailed output. Additionally,
by using a separate particle filter for each voxel's time
series it should be feasible to calculate parameters and make real-time predictions
for small neural regions, a feature which could be useful towards real time FMRI 
(\cite{DeCharms2005}).

For the past twenty years Functional Magnetic Resonance Imaging (FMRI) 
has been at the forefront of cognitive research. Despite it's
limited temporal resolution, FMRI is the standard tool for localizing 
neural activation.  Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is quick and cheap, and its analysis straight forward.
By modeling the governing equations behind the neural response that
drives FMRI, it is possible to increase the power of FMRI.
The underlying state equations hold important information
about how individual brain regions react to stimuli. The model parameters
on the other hand, hold important information about the patients individual
physiology including existing and future pathologies. In short,
the long chain of events driving FMRI signals contain information 
beyond correlation with stimuli.

@@

\section{Post Stimulus Undershoot}
\label{sec:Post Stimulus Undershoot}
Although the most widely used, the BOLD model described in \autoref{eq:bold}
and \autoref{eq:boldout} have been extensively added on to. The most
significant feature missing from the original model is the 
"post-stimulus undershoot".
The "post-stimulus" undershoot is the name for a prolonged subnormal
BOLD response for a period of 10 to 60 seconds after stimulus has
ceased (\cite{Chen2009}, \cite{Mandeville1999a}).

@@

\section{Noise}
\label{sec:Introduction Noise}
As demonstrated in \autoref{sec:BOLD Physiology} the BOLD response has been
extensively studied and despite minor discrepancies, the cause of the BOLD 
signal is well known. However, as FMRI detects an  
aggregate signal over the space of cubic centimeters, there are
plenty of sources of noise. Though local neurons act
"together" (i.e. around the same time), the density of neurons, the
density of capillaries, and slight differences in activation across 
a particular voxel can all lead to signal attenuation and noise.

A particularly difficult form of noise present in FMRI is a low frequency
drift, often characterized as a Wiener process (\cite{Riera2004}). 
Though not present in all regions, as much as ten to fifteen percent
of voxels can be affected (\cite{Tanabe2002}), thus it is prevalent enough to cause significant
inference problems \cite{Smith2007}. It is still not
clear what exactly causes this noise noise comes from, although it is possible it is 
the result of magnets heating up, or some distortion in magnetic
fields \cite{Smith2007}. It is clear that this drift signal is not solely
due to a physiological effects, given its presence in cadavers and phantoms 
\cite{Smith1999}. Interestingly, it is usually spatially correlated, and
more prevalent at interfaces between regions, though 
by no means limited to such areas. Though one potential source
could be slight movement, given that co-registration of volumes to a single time
point is standard, this seems unlikely. Regardless, the problem mandates
the use of a high pass filter to make inference reasonably powered
\cite{Smith2007}.

In order to characterize the noise, I analyzed resting state data.
During resting state, the patient is shown no images, and he is asked
to avoid movement and complex thought, to the best of his abilities.
Some limitations exist, for instance EPI sequences are extremely loud
which could cause some stimulus in the auditory cortex, and of course
very long sequences could give the patient time for his mind to wander.
Overall though there should be no activation, and thus the signal will
consist entirely of noise. Therefore resting state data is perfect
for analyzing the noise distribution. The locations were chosen from
various points around the brain, all in grey matter voxels. The time
series were also chosen because they were representative of different types
of noise I found in the resting state data.

Because most methods (including the one used in this paper)
assume the noise realizations are independent of each other, the auto-
correlation is of particular interest (which is a necessary but not
sufficient condition for independence). Gaussianity is also a common
assumption made in studies of FMRI data, though that assumption is not
made in this work. Regardless, comparing the distribution to a Gaussian
is informative, so I used Q-Q plots to compare example data with the
Normal distribution. Additionally, in FMRI data the noise is often considered 
to be Wiener \cite{Riera2003}. Recall that a Wiener random process is
characterized by steps that are Gaussian; in other words the difference between
any two samples is Normally distributed. The simulations discussed in 
\autoref{sec:Single Voxel Simulation} make use of this, 
by adding a Wiener random process to the overall signal. To determine
whether the noise is in fact Wiener, the distribution of 
the difference between adjacent measurements was plotted against
a Gaussian. Wiener processes further assume that the steps are independent;
so the steps should have near flat autocorrelation functions.

Finally, removal of the so called "drift" is often
performed with some variation of a high pass filter, 
so I checked the noise distribution after applying such a filter (in
this case the subtraction of a spline, see \autoref{sec:Methods Preprocessing}).
Here I wanted to know how effective my high pass filter at the removal of
Wiener noise.

\autoref{fig:QQDC} shows the 
results with a regression line fit to the points on the Q-Q plot.
Recall that in a Quartile-Quartile (Q-Q) plot, if the points plotted on the 
x-axis and the points
on the y-axis come from the same \emph{type} of distribution then all the points will
fall on a line. Differences in the variance will cause the line to have a slope
other than 1, while differences in the expected value will cause the fitted line
to be shifted. In these Q-Q plots, the points are being compared to the standard
Gaussian distribution, so the quality of the line fit determines how closely the points
fit the Gaussian distribution. Note that in \autoref{fig:QQDC} the points have all been 
normalized (changed to percent difference).
\begin{figure}
\centering
\subfigure[]{\label{fig:QQDC:A}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009_29_49_9}}
\subfigure[]{\label{fig:QQDC:B}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009_34_43_24}}
\subfigure[]{\label{fig:QQDC:C}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009_22_38_23}}
\subfigure[]{\label{fig:QQDC:D}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009_37_29_24}}

%\subfigure{\includegraphics[trim=6cm 1cm 0 0cm,width=17cm]{images/noise_0009_19-24-10.pdf}}
%\subfigure{\includegraphics[trim=6cm 1cm 0 0cm,width=17cm]{images/noise_0009_20-45-18.pdf}}
%\subfigure{\includegraphics[trim=6cm 1cm 0 0cm,width=17cm]{images/noise_0009_23-47-18.pdf}}
%\subfigure{\includegraphics[trim=6cm 1cm 0 0cm,width=17cm]{images/noise_0009_35-49-9.pdf}}

\caption{Q-Q Plots of normalized resting state data}
\label{fig:QQDC}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{fig:QQDDelta:A}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_29_49_9}}
\subfigure[]{\label{fig:QQDDelta:B}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_34_43_24}}
\subfigure[]{\label{fig:QQDDelta:C}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_22_38_23}}
\subfigure[]{\label{fig:QQDDelta:D}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_37_29_24}}
\caption{Q-Q Plots of resting state data, using the BOLD signal changes}
\label{fig:QQDelta}
\end{figure}

A Wiener process should still conform to the normal distribution, with a 
variance proportional to the run-time. Note that \autoref{fig:QQDC:A} and \autoref{fig:QQDC:B}
are relatively well described by a Gaussian process with a small autocorrelation, 
\autoref{fig:QQDC:C} and \autoref{fig:QQDC:D} are not. In particular the tails of \autoref{fig:QQDC:C}
do not seem to fit the Gaussian well. Also note the significant autocorrelation in
\autoref{fig:QQDC:C} and \autoref{fig:QQDC:D}. As expected, the noise is not strictly
Gaussian white noise.  On the other hand, the steps do conform rather
closely to the normal distribution.
As expected, most of the autocorrelation disappears for the step data. Given
that the steps seem to fit the Normal distribution, the low autocorrelation
indicates that the steps could be Independently Distributed. 
Therefore, the assumption that the noise follows a Wiener process seems to
be correct.

\begin{figure}
\centering
\subfigure[]{\label{fig:QQs:A}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009s_29_49_9}}
\subfigure[]{\label{fig:QQs:B}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009s_34_43_24}}
\subfigure[]{\label{fig:QQs:C}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009s_22_38_23}}
\subfigure[]{\label{fig:QQs:D}\includegraphics[trim=6cm 1cm 6cm 0cm,width=14cm]{images/noise2_0009s_37_29_24}}
\caption{Q-Q Plots of resting state data, after the de-trending}
\label{fig:QQSpline}
\end{figure}

De-trending the time-series by subtracting a spline fit to the distribution
removed much of the autocorrelation present in \autoref{fig:QQDC:C} and \autoref{fig:QQDC:D},
though clearly not perfectly. Though the distributions still do not exactly fit
the Normal, \autoref{fig:QQs:D} clearly is much improved compared to \autoref{fig:QQDC:D}.
In all, the de-trending is effectively removing Wiener noise.

>>>>

\chapter{Introduction}
Traditional methods of analyzing timeseries images produced by 
Function Magnetic Resonance 
Imaging (FMRI) use regression techniques with very few degrees of freedom. 
Though adding more degrees of freedom naturally mandates more computation,
in this thesis I will discuss a Sequential Monte Carlo method of fitting a nonlinear
model with seven degrees of freedom at 
a computation cost that would still allow real time calculations for multiple voxels.
More practically, the method described here forms a completely
separate method of detecting neural activity in comparison to the traditionally
used Statistical Parametric Mapping (SPM). Though more computationally intense,
this method is capable of modeling nonlinear effects, is conceptually simpler
and provides more detailed output. Additionally,
by using a separate particle filter for each voxel's time
series it is possible to estimate parameters and make real-time predictions
for small neural regions, a feature which could be useful towards real time FMRI 
(\cite{DeCharms2005}). Future works will also benefit from the ability to 
apply conditions to the posterior distribution in post-processing without
having to re-run the algorithm; for instance in light of additional physiological
constraints. Modeling the BOLD response as a nonlinear system is the
best way to determine the correlation of stimulus sequence with the BOLD
response; yet in the past doing this on a large scale has been far too
computationally taxing. The solution used here takes approximately 40 seconds
for a single voxel's time series (5 minutes in length, with a quad core machine).

This thesis is organized as follows. In the introduction I will introduce
FMRI, the method by which neural time changing data is detected. This section
will also describe the basic form of the BOLD model - which drives the 
detectable changes in MR signal. \autoref{sec:Prior Works} will discuss other
methods of analyzing FMRI images as well as other techniques that have
been, or could be applied to the nonlinear regression model described here. 
\autoref{sec:Particle Filter} derives the particle filter using Bayesian 
statistics and discusses some practical elements of implementing the 
particle filter algorithm. \autoref{sec:Methods} then goes into further
detail about the specific particle filter configuration used in this work.
This section also describes the pre-processing used before the particle
filter is applied. The results are described separately for simulated data
and real FMRI data in \autoref{sec:SimulationResults} and \autoref{sec:RealData},
respectively. Finally in \autoref{sec:Conclusion} there is a discussion of
the usefulness and implications of this technique as well as recommendations
for direction of future works.

\section{Historic Context}
For the past twenty years, Functional Magnetic Resonance Imaging (FMRI) 
has been at the forefront of cognitive research. Despite it's
limited temporal resolution, FMRI is the standard tool for localizing 
neural activation.  Whereas other methods
of analyzing neural signals can be invasive or difficult to acquire, 
FMRI is quick and cheap, and its analysis straight forward.
By modeling the governing equations behind the neural response that
drives FMRI, it is possible to increase the power of FMRI.
The underlying state equations hold important information
about how individual brain regions react to stimuli. The model parameters
on the other hand, hold important information about the patients individual
physiology including existing and future pathologies. In short,
the long chain of events driving FMRI signals contain information 
beyond correlation with stimuli.

@@

\section{Post Stimulus Undershoot}
\label{sec:Post Stimulus Undershoot}
Although the most widely used, the BOLD model described in \autoref{eq:bold}
and \autoref{eq:boldout} have been extensively added on to. The most
significant feature missing from the original model is the 
post-stimulus undershoot.
The post-stimulus undershoot is the term used for a prolonged subnormal
BOLD response for a period of 10 to 60 seconds after stimulus has
ceased (\cite{Chen2009}, \cite{Mandeville1999a}).

@@
